---
title: "Using the LibGEOS Rcpp API"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

The [GEOS](https://trac.osgeo.org/geos/) library is written in C++, with its recommended interface as a [C API](https://trac.osgeo.org/geos/browser/git/capi/geos_c.h.in) for long-term stability. Because of this long-term stability (and because there is no other way to do cross-package dynamic linking in R), this is the interface exposed via the libgeos package. While it seems a bit silly to have a massively powerful and well-written C++ library squeezed through the tiny hole of the C API, it powers a [huge number of libraries and products](https://trac.osgeo.org/geos/wiki/Applications), arguably because of the stability of the C API.

Arguably even more silly is to write a C++ API wrapping the C API (that was written around a C++ library in the first place). However, in R, header-only C++ libraries are easy to distribute (a well-known example is [Rcpp](https://cran.r-project.org/package=Rcpp)), and make it easier for more people to write safe high-performance code. Rcpp isn't the only way to use C++ to interact with R, but it is the most popular, and there are excellent [reference guides](https://teuder.github.io/rcpp4everyone_en/) and introductory material for [scripting](https://adv-r.hadley.nz/rcpp.html) and [using it in packages](http://r-pkgs.had.co.nz/src.html).

## Getting Started

To get started, you'll have to install and load libgeos:

```{r setup}
library(libgeos)
```

Next, you'll need to make a new `.cpp` file and use `Rcpp::sourceRcpp()` to make the exported functions available. The anatomy of a file that uses the libgeos Rcpp API is as follows:

```{Rcpp}
#include <Rcpp.h>
// [[Rcpp::depends(libgeos)]]
#include "libgeos-rcpp.h"
#include "libgeos.c"

// [[Rcpp::export]]
void cpp_libgeos_init_api() {
  libgeos_init_api();
}

// regular C or C++ code that uses GEOS functions!
// [[Rcpp::export]]
std::string version() {
  return GEOSversion();
}
```

A few things to note:

- Use `// [[Rcpp::depends(libgeos)]]` to tell Rcpp to put libgeos's headers in the include path
- Use `#include "libgeos-rcpp.h"` to include the GEOS function definitions and the Rcpp API
- Use `#include "libgeos.c"` to include the actual GEOS function pointers.
- You need to provide a way to call `libgeos_init_api()` to initialize the function pointers with references to the implementations in libgeos.

After compiling the above with `sourceRcpp()`, in R you can do:

```{r}
cpp_libgeos_init_api() # once
version()
```

In a package this will be slightly different: you need to include `libgeos.c` exactly once, but all other files that use GEOS functions only need to `#include "libgeos-rcpp.h"`. You don't need `// [[Rcpp::depends(libgeos)]]` in any of your package files, but you will need to add `LinkingTo: libgeos` to your `DESCRIPTION` and make sure to call `libgeos_init_api()` when your package is loaded (e.g., via `.onLoad()`).

## The LibGEOSHandle

Every function in the GEOS API exported by libgeos (except `GEOSVersion()`) requires a `GEOSContextHandle_t` object as the first argument. This object primarily controls error reporting and makes the API thread-safe if used correctly. The libgeos Rcpp API wraps this object with the `LibGEOSHandle` class, which takes care of creating and destroying the underlying `GEOSContextHandle_t`, and reporting errors in such a way that R doesn't crash.

To use the `LibGEOSHandle`, declare it at the beginning of your function and pass it everywhere. Classes that are part of the libgeos Rcpp API take the `LibGEOSHandle` object, but GEOS API functions need the underlying `GEOSContextHandle_t` (available via `.get()`).

```{Rcpp, eval = FALSE}
LogicalVector wkt_is_empty(CharacterVector wkt) {
  LibGEOSHandle handle;
  LibGEOSWKTReader reader(handle);
  
  LogicalVector output(wkt.size());
  for (R_xlen_t i = 0; i < wkt.size(); i++) {
    LibGEOSGeometry geom = reader.read(wkt[i]);
    output[i] = GEOSisEmpty_r(handle.get(), geom.get());
  }
  
  return output;
}
```

One thing that is important to note is that raw `GEOS*_r()` functions never throw exceptions. This is by design: throwing an exception across a library boundary doesn't always work. In the above example, you could check the return value of `GEOSisEmpty_r()` and throw the corresponding exception if one occurred. The `LibGEOSHandle` collects the error messages reported by GEOS, and you can throw the last one it encountered using `handle.throwLastError()` (or do something else with the message using `handle.lastErrorMessage()`). You will have to [consult the GEOS C API header](https://trac.osgeo.org/geos/browser/git/capi/geos_c.h.in) to see what return value corresponds to which error codes. Convenience methods `handle.checkGeometry()` and `handle.checkPredicateResult()` are provided for these two common cases (geometry generating functions return `NULL` on error; predicates return `2` on error).

Note that `LibGEOS*` objects do throw exceptions, so you only have to worry about this when you call the GEOS C API functions yourself.

## Reading and Writing Geometry

GEOS has two main ways of getting geometry in and out: well-known binary and well-known text. In R, these correspond to `List`s of `RawVector`s and `CharacterVector`s, respectively. You can use the `LibGEOSWKBReader` and `LibGEOSWKTReader` to read geometries, and `LibGEOSWKBWriter` and `LibGEOSWKTWriter` to write them. I tend to use WKT for examples, but WKB is orders of magnitude faster (and you can generate it readily from [sf](https://r-spatial.github.io/sf/) objects using `st_as_binary()`). As an example, the following function reads WKT and writes it back again with the settings that you probably want (notably, `writer.setTrim(true)`).

```{Rcpp}
#include <Rcpp.h>
// [[Rcpp::depends(libgeos)]]
#include "libgeos-rcpp.h"
#include "libgeos.c"
using namespace Rcpp;

// [[Rcpp::export]]
void cpp_libgeos_init_api() {
  libgeos_init_api();
}

// [[Rcpp::export]]
CharacterVector wkt_convert_wkt(CharacterVector wkt) {
  LibGEOSHandle handle;
  LibGEOSWKTReader reader(handle);
  LibGEOSWKTWriter writer(handle);
  writer.setTrim(true);
  
  CharacterVector output(wkt.size());
  for (R_xlen_t i = 0; i < wkt.size(); i++) {
    LibGEOSGeometry geom = reader.read(wkt[i]);
    output[i] = writer.write(geom.get());
  }

  return output;
}
```

```{r}
cpp_libgeos_init_api()
wkt_convert_wkt("POINT (-64 45)")
```

Note that the `.read()` functions always return a `LibGEOSGeometry`, which wraps the `GEOSGeometry*` pointer to make sure that the object is destroyed when the variable goes out of scope. You can get a reference to the `GEOSGeometry*` using `.get()` to pass it to the GEOS C API functions directly.

## Predicates

Geometry predicates in GEOS return 0 for false, 1 for true, or 2 if an exception occurred. Most predicates in GEOS operate on two geometries (e.g., `GEOSIntersects_r()`), but a few operate on one (e.g., `GEOSisEmpty_r()`). Finally, there are the "prepared" predicates, which build a spatial index on the geometry to more efficiently compute (potentially many) relationships. Because the lifecycles of the `GEOSGeometry*` and the `GEOSPreparedGeometry*` are intertwined, these are both handled by the `LibGEOSGeometry` class (the geometry is prepared the first time `.getPrepared()` is called). Examples of each of these types of predicates are provided below.

```{Rcpp}
#include <Rcpp.h>
// [[Rcpp::depends(libgeos)]]
#include "libgeos-rcpp.h"
#include "libgeos.c"
using namespace Rcpp;

// [[Rcpp::export]]
void cpp_libgeos_init_api() {
  libgeos_init_api();
}

// [[Rcpp::export]]
LogicalVector wkt_is_empty(CharacterVector wkt) {
  LibGEOSHandle handle;
  LibGEOSWKTReader reader(handle);
  
  LogicalVector output(wkt.size());
  for (R_xlen_t i = 0; i < wkt.size(); i++) {
    LibGEOSGeometry geom = reader.read(wkt[i]);
    char result = GEOSisEmpty_r(handle.get(), geom.get());
    output[i] = handle.checkPredicateResult(result);
  }

  return output;
}

// [[Rcpp::export]]
LogicalVector wkt_intersects(CharacterVector x, CharacterVector y) {
  LibGEOSHandle handle;
  LibGEOSWKTReader reader(handle);
  
  LogicalVector output(x.size());
  for (R_xlen_t i = 0; i < x.size(); i++) {
    LibGEOSGeometry geomX = reader.read(x[i]);
    LibGEOSGeometry geomY = reader.read(y[i]);
    char result = GEOSIntersects_r(handle.get(), geomX.get(), geomY.get());
    output[i] = handle.checkPredicateResult(result);
  }

  return output;
}

// [[Rcpp::export]]
LogicalVector wkt_intersects_prepared(CharacterVector x, CharacterVector y) {
  LibGEOSHandle handle;
  LibGEOSWKTReader reader(handle);
  
  LogicalVector output(x.size());
  for (R_xlen_t i = 0; i < x.size(); i++) {
    LibGEOSGeometry geomX = reader.read(x[i]);
    LibGEOSGeometry geomY = reader.read(y[i]);
    char result = GEOSPreparedIntersects_r(handle.get(), geomX.getPrepared(), geomY.get());
    output[i] = handle.checkPredicateResult(result);
  }

  return output;
}
```

```{r}
cpp_libgeos_init_api()
wkt_is_empty(c("POINT (-64 45)", "POINT EMPTY"))
wkt_intersects("POINT (0 0)", "POINT (0 0)")
wkt_intersects_prepared("POINT (0 0)", "POINT (0 0)")
```

## Transformers

GEOS Functions that return a geometry add a small layer of complexity in that they create a `GEOSGeometry*` that must be cleaned up afterward. You can use `LibGEOSGeometry` to manage these objects, as demonstrated by the following wrapper of `GEOSGetCentroid_r()`. The constructor of `LibGEOSGeometry()` will throw an exception with an  informative error if it's passed `NULL`, which almost always denotes an error for these operations.

```{Rcpp}
#include <Rcpp.h>
// [[Rcpp::depends(libgeos)]]
#include "libgeos-rcpp.h"
#include "libgeos.c"
using namespace Rcpp;

// [[Rcpp::export]]
void cpp_libgeos_init_api() {
  libgeos_init_api();
}

// [[Rcpp::export]]
CharacterVector wkt_centroid(CharacterVector wkt) {
  LibGEOSHandle handle;
  LibGEOSWKTReader reader(handle);
  LibGEOSWKTWriter writer(handle);
  writer.setTrim(true);
  
  CharacterVector output(wkt.size());
  for (R_xlen_t i = 0; i < wkt.size(); i++) {
    LibGEOSGeometry geom = reader.read(wkt[i]);
    LibGEOSGeometry result(handle, GEOSGetCentroid_r(handle.get(), geom.get()));
    output[i] = writer.write(result.get());
  }

  return output;
}
```

```{r}
cpp_libgeos_init_api()
wkt_centroid(c("LINESTRING (0 0, 10 10)", "LINESTRING EMPTY"))
```

## Extractors

An important note about many functions that compute a `double` is that they often pass these values to the caller using a pointer argument (so that they can `return` an error code). An example is `GEOSArea_r()`, which returns an error code on failure and passes its area argument on via a pointer. Wrappers that use these functions usually look like this:

```{Rcpp, eval = FALSE}
NumericVector wkt_is_empty(CharacterVector wkt) {
  LibGEOSHandle handle;
  LibGEOSWKTReader reader(handle);
  
  NumericVector output(wkt.size());
  for (R_xlen_t i = 0; i < wkt.size(); i++) {
    LibGEOSGeometry geom = reader.read(wkt[i]);
    double area;
    GEOSArea_r(handle.get(), geom.get(), &area);
    output[i] = area;
  }
  
  return output;
}
```

## Using the C API

If you don't need the features of C++, you can import only the C API via `libgeos.h`. You will also have to include `libgeos.c` exactly once in your shared library and call `libgeos_init_api()` before using any GEOS functions.
